---
title: "Methylation_Processing_CNA"
author: "Ruchit Patel"
date: "2023-09-23"
output: html_document
---

---
This script uses "minifi", "conumee", and "IlluminaHumanMethylationEPICanno.ilm10b4.hg19" from Bioconductor. All contributions to Bioconductor exist under an open source license. "minfi" is under Artistic-2.0 License, "conumee" under GPL (>=2) License, and "IlluminaHumanMethylationEPICanno.ilm10b4.hg19" under Artistic-2.0 License. "minfi" citation is Kasper et al., "conumee" citation is 
Volker Hovestadt et al., and "IlluminaHumanMethylationEPICanno.ilm10b4.hg19" citation is Kasper et al.
---


```{r}
library(minfi)
library(conumee)
library(dplyr)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)

#Load meningioma data. This will help process the multifocal sampled data and the primary-recurrent external validation samples that are also methylation dta. Datafiles are from GSE151067 on NCBI GEO (multi-focal sampled data) and GSE228316 (primary-recurrent samples)

RGmening <- read.metharray.exp('Path to file here') #/Downloads/Multifocal_Raw or /Downloads/Prim_Rec_Lucas_Raw
Methylmening <- preprocessIllumina(RGmening)

mening_data <- CNV.load(Methylmening)

#Load reference data - whole blood
RGblood <- read.metharray.exp('Path to whole blood reference data') #Path to whole blood reference data from GSE235717 on NCBI GEO (whole blood)
Methylblood <- preprocessIllumina(RGblood)

blood_data <- CNV.load(Methylblood)

#Create annotation file for probes
mening_anno <- CNV.create_anno(array_type = "EPIC")

mening_anno@probes <- mening_anno@probes[names(mening_anno@probes) %in% names(minfi::getLocations(IlluminaHumanMethylationEPICanno.ilm10b4.hg19::IlluminaHumanMethylationEPICanno.ilm10b4.hg19))]


#Loop through samples and create probe info

mening_names <- names(mening_data)
combined_seg_means <- data.frame()
counter <- 0

# Loop through each name in mening_names
for (sample_name in mening_names) {
  #Increment the counter
  counter <- counter + 1
  
  # Update the sample name in your code
  x <- CNV.fit(mening_data[sample_name], blood_data, mening_anno)
  x <- CNV.bin(x)
  x <- CNV.segment(x)
  
  #CNV.genomeplot(x)
  
  # Use CNV.write to output to a file with the sample name
  seg_means <- CNV.write(x, what = "segments")
  
  # Append the results to the combined_seg_means data frame
  combined_seg_means <- rbind(combined_seg_means, seg_means)
  
  # Print the progress
  cat("Processing sample", counter, "of", length(mening_names), "\n")
}

# Save the combined results to a single file
write.table(combined_seg_means, file = "Prim_rec_lucas_seg_means.txt", sep = "\t", quote = FALSE, row.names = FALSE) #Multifocal_mening_seg_means.txt
```


```{r}
library(RColorBrewer)
library(ggplot2)
library(ggridges)
library(forcats)
library(fmsb)
library(forestplot)
library(dplyr)
library(openxlsx)
library(stringr)
library(scales)
library(tidyr)
library(gridExtra)
library(cowplot)
library(pracma)
library(networkD3)
library(survival)
library(broom)
library(survminer)

path <- 'Path to data file output from above'
file_name <- "Multifocal_mening_seg_means.txt" #Prim_rec_lucas_seg_means.txt or Multifocal_mening_seg_means.txt, datafiles from above
genome_file_name <- 'genome_reference_GRCh37.xlsx' #genome_reference_GRCh38.p14.xlsx
arm_file_name <- 'arm_reference_GRCh37.xlsx' #arm_reference_GRCh38.p14.xlsx
file_path <- file.path(path, file_name)
genome_reference_file_path <- file.path(path, genome_file_name)
arm_reference_file_path <- file.path(path, arm_file_name)

#Import meningioma master seg file and genome size file
multifoc_mening_seg <- read.table(file_path, sep = "\t", header = TRUE, stringsAsFactors = FALSE)
genome_size <- read.xlsx(genome_reference_file_path)
arm_size <- read.xlsx(arm_reference_file_path)

```

```{r}
#Isolate sample identifiers and locations: this is for multifocal samples

multifoc_mening_seg <- multifoc_mening_seg %>%
  mutate(
    sample = sub("_.*$", "", ID)
  ) %>%
  select(sample, everything())

#Removing rows corresponding to sample that was a liver met, not part of cranial meningiomas
multifoc_mening_seg <- multifoc_mening_seg %>% filter(sample != "GSM4565801")


multifoc_id_file_name <- 'multifoc_id.xlsx'
multifoc_id_file_path <- file.path(path, multifoc_id_file_name)
multifoc_id <- read.xlsx(multifoc_id_file_path)

multifoc_mening_seg <- multifoc_mening_seg %>% left_join(multifoc_id, by = "sample")

multifoc_mening_seg <- multifoc_mening_seg %>% select(-ID, -individual, -location)

#old code

multifoc_mening_seg <- multifoc_mening_seg %>%
  mutate(
    sample = sub("^[^_]*_(.*?)_.*$", "\\1", ID),
    anat_loc = sub("^[^_]*_[^_]*_(.*)$", "\\1", ID)
  ) %>%
  select(sample, anat_loc, everything())

# Create a data frame with unique values and identifiers
unique_samples <- unique(multifoc_mening_seg$sample)
identifiers <- paste0("MM", seq_len(length(unique_samples)))
sample_identifier_df <- data.frame(sample = unique_samples, identifier = identifiers)

# Merge the identifier data with the original data
multifoc_mening_seg <- multifoc_mening_seg %>%
  left_join(sample_identifier_df, by = "sample")

multifoc_mening_seg <- multifoc_mening_seg %>%
  mutate(
    digit_from_anat_loc = sub(".*R0(\\d+)C\\d+", "\\1", anat_loc),
    identifier = paste0(identifier, "_", digit_from_anat_loc)
  ) %>%
  select(-digit_from_anat_loc)  # Remove the intermediate column if needed

multifoc_mening_seg <- multifoc_mening_seg %>%
  select(identifier, everything(), -c(sample, anat_loc, ID))
```

```{r}
#For primary/recurrent analysis, isolate sample identifiers and merge in sample names: proceed with the next code chunk if you run this

multifoc_mening_seg <- multifoc_mening_seg %>%
  mutate(
    sample = sub("_.*$", "", ID)
  ) %>%
  select(sample, everything())

prim_rec_id_file_name <- 'primary_recurrent_external_lucas.xlsx'
prim_rec_id_file_path <- file.path(path, prim_rec_id_file_name)
prim_rec_id <- read.xlsx(prim_rec_id_file_path)

multifoc_mening_seg <- multifoc_mening_seg %>%
  left_join(prim_rec_id, by = c("sample" = "GEO.ID")) %>%
  mutate(identifier = Sample.ID) %>%
  select(-Sample.ID)

```

```{r}
#Compute probe location changes
multifoc_mening <- multifoc_mening_seg %>% filter(!(seg.mean >= -0.1 & seg.mean <= 0.15))

#Annotate loss and gain events
multifoc_mening <- multifoc_mening %>% mutate(event = ifelse(seg.mean < 0, "loss", "gain"))

multifoc_mening <- multifoc_mening %>% rename(Chr = chrom)

multifoc_mening <- merge(multifoc_mening, genome_size, by = "Chr", all.x = TRUE)

multifoc_mening <- multifoc_mening %>% arrange(identifier)

#Figure out which arms are affected

multifoc_mening <- multifoc_mening %>%
  mutate(Arm = case_when(
    loc.start < arms_division & loc.end < arms_division ~ "p",
    loc.start > arms_division & loc.end > arms_division ~ "q",
    loc.start < arms_division & loc.end > arms_division ~ "pq",
    TRUE ~ NA_character_  # Handle other cases (optional)
  ))

multifoc_mening <- multifoc_mening %>% mutate(arm_individual = Arm)

# Duplicate rows where arm_individual equals "pq" and set one row with "p"
duplicated_rows_p <- multifoc_mening %>%
  filter(arm_individual == "pq") %>%
  mutate(arm_individual = "p")

# Duplicate rows where arm_individual equals "pq" and set one row with "q"
duplicated_rows_q <- multifoc_mening %>%
  filter(arm_individual == "pq") %>%
  mutate(arm_individual = "q")

# Remove the original rows where arm_individual equals "pq"
multifoc_mening <- multifoc_mening %>%
  filter(arm_individual != "pq")

# Combine the original dataframe and duplicated rows with "p" and "q"
multifoc_mening <- bind_rows(multifoc_mening, duplicated_rows_p, duplicated_rows_q)

multifoc_mening <- multifoc_mening %>%
  mutate(size_affected = case_when(
    Arm == "p" ~ loc.end - loc.start,
    Arm == "q" ~ loc.end - loc.start,
    Arm == "pq" & arm_individual == "p" ~ arms_division - loc.start,
    Arm == "pq" & arm_individual == "q" ~ loc.end - arms_division,
  ))

#Calculate percent arm affected
multifoc_mening <- merge(multifoc_mening, arm_size[, c("Chr", "Arm", "size_arm")], 
                         by.x = c("Chr", "arm_individual"), by.y = c("Chr", "Arm"), 
                         all.x = TRUE)

multifoc_mening <- multifoc_mening %>%
  mutate(pct_affected = (size_affected / size_arm) * 100)

```


```{r}
#Create separate frames for gains and losses, noncontiguous loss (combine it all together)

multifoc_noncont_loss <- multifoc_mening %>%
  filter(event == "loss") %>%
  select(Chr, arm_individual, identifier, pct_affected)

multifoc_noncont_gain <- multifoc_mening %>%
  filter(event == "gain") %>%
  select(Chr, arm_individual, identifier, pct_affected)

multifoc_noncont_loss <- multifoc_noncont_loss %>%
  group_by(Chr, arm_individual, identifier) %>%
  summarize(pct_affected = sum(pct_affected))

multifoc_noncont_gain <- multifoc_noncont_gain %>%
  group_by(Chr, arm_individual, identifier) %>%
  summarize(pct_affected = sum(pct_affected))

#Stop here if analyzing primary/recurrent data and move on to few code chunks down. This needs to be processed differently for plotting like the other primary/recurrent samples.

multifoc_noncont_loss <- multifoc_noncont_loss %>%
  separate(identifier, into = c("identifier", "sample_num"), sep = "_")

multifoc_noncont_gain <- multifoc_noncont_gain %>%
  separate(identifier, into = c("identifier", "sample_num"), sep = "_")

multifoc_noncont_loss <- multifoc_noncont_loss %>%
  mutate(pct_affected = ifelse(pct_affected > 100, 100, pct_affected))

multifoc_noncont_gain <- multifoc_noncont_gain %>%
  mutate(pct_affected = ifelse(pct_affected > 100, 100, pct_affected))
```

```{r}
#Analysis of loss
MM6_loss <- multifoc_noncont_loss %>% filter(identifier == "MM6")

MM6_loss$sample_num <- as.integer(MM6_loss$sample_num)

# Get unique combinations of Chr and arm_individual from the MM6_loss dataframe
existing_combinations <- MM6_loss %>%
  select(Chr, arm_individual) %>%
  distinct()

# Create a dataframe with all possible sample_num values from 1 to 8 (adjust this based on number of times we have sampled)
sample_nums <- data.frame(sample_num = 1:7)

# Cross join the existing combinations with sample_nums
MM6_loss <- existing_combinations %>%
  merge(sample_nums, by = NULL) %>%
  left_join(MM6_loss, by = c("Chr", "arm_individual", "sample_num"))

# Fill missing values in pct_affected with 0
MM6_loss$pct_affected[is.na(MM6_loss$pct_affected)] <- 0

MM6_loss <- MM6_loss %>%
  arrange(Chr, arm_individual, sample_num)


#Analysis of gains
MM6_gain <- multifoc_noncont_gain %>% filter(identifier == "MM6")

# Convert the "sample_num" column to integer
MM6_gain$sample_num <- as.integer(MM6_gain$sample_num)

# Get unique combinations of Chr and arm_individual from the MM6_gain dataframe
existing_combinations <- MM6_gain %>%
  select(Chr, arm_individual) %>%
  distinct()

# Create a dataframe with all possible sample_num values from 1 to 8 (adjust this based on number of times we have sampled)
sample_nums <- data.frame(sample_num = 1:7)

# Cross join the existing combinations with sample_nums
MM6_gain <- existing_combinations %>%
  merge(sample_nums, by = NULL) %>%
  left_join(MM6_gain, by = c("Chr", "arm_individual", "sample_num"))

# Fill missing values in pct_affected with 0
MM6_gain$pct_affected[is.na(MM6_gain$pct_affected)] <- 0

MM6_gain <- MM6_gain %>%
  arrange(Chr, arm_individual, sample_num)
```

```{r}
#Plotting values

MM6_loss$chr_arm <- paste(MM6_loss$Chr, MM6_loss$arm_individual, sep = "_")

MM6_loss_plot <- ggplot(MM6_loss, aes(x = chr_arm, y = pct_affected)) +
  geom_jitter(position = position_jitter(0.2), size = 4, color = "deepskyblue3") +
  stat_summary(fun.y = median, geom = "crossbar", color = "gray52", size = 0.35, width = 0.5) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.ticks = element_blank(),
    axis.text = element_text(size = 18)
  ) + scale_y_continuous(breaks = seq(0, 100, by = 10), limits=c(-1,101))


MM6_gain$chr_arm <- paste(MM6_gain$Chr, MM6_gain$arm_individual, sep = "_")

MM6_gain_plot <- ggplot(MM6_gain, aes(x = chr_arm, y = pct_affected)) +
  geom_jitter(position = position_jitter(0.2), size = 4, color = "firebrick3") +
  stat_summary(fun.y = median, geom = "crossbar", color = "gray52", size = 0.35, width = 0.5) +
  scale_y_continuous(breaks = seq(0, 100, by = 10), limits = c(-1,101)) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.ticks = element_blank(),
    axis.text = element_text(size = 18)
  )

```


```{r}
#Handling methylation based primary-recurrent samples

# Check if 'identifier' ends with '_R' and assign 0 or 1 to recurrent or primary samples
multifoc_noncont_loss <- multifoc_noncont_loss %>% mutate( primary = if_else(grepl("_R$", identifier), 0, 1))
multifoc_noncont_gain <- multifoc_noncont_gain %>% mutate( primary = if_else(grepl("_R$", identifier), 0, 1))

#Reframe identifier column to prepare for plotting
multifoc_noncont_loss <- multifoc_noncont_loss %>% rename(old_ID = identifier)
multifoc_noncont_gain <- multifoc_noncont_gain %>% rename(old_ID = identifier)

multifoc_noncont_loss <- multifoc_noncont_loss %>% mutate(identifier = sub("_R$", "", old_ID))
multifoc_noncont_gain <- multifoc_noncont_gain %>% mutate(identifier = sub("_R$", "", old_ID))

id_mapping <- c("M4" = "PR1", "M5" = "PR2", "M6" = "PR3")

multifoc_noncont_loss <- multifoc_noncont_loss %>% mutate(identifier = case_when(identifier %in% names(id_mapping) ~ id_mapping[identifier], TRUE ~ identifier))
multifoc_noncont_gain <- multifoc_noncont_gain %>% mutate(identifier = case_when(identifier %in% names(id_mapping) ~ id_mapping[identifier], TRUE ~ identifier))

methyl_pr_loss <- multifoc_noncont_loss
methyl_pr_gain <- multifoc_noncont_gain

methyl_pr_loss <- methyl_pr_loss %>% mutate(Chr_Arm = paste0(sub("chr", "", Chr), arm_individual))
methyl_pr_gain <- methyl_pr_gain %>% mutate(Chr_Arm = paste0(sub("chr", "", Chr), arm_individual))
```

```{r}
#Setting up primary/recurrent analysis
#Ensure that any 0 values are plotted (e.g., should be pairs of dots for primary/recurrent)

methyl_pr_loss_mod <- data.frame()

# Loop through identifier values from PR1 to PR13
for (i in 1:3) {
  identifier_value <- paste0("PR", i)
  
  # Filter the data for the current identifier
  subset_data <- methyl_pr_loss %>% filter(identifier == identifier_value)
  
  # Find unique Chr_Arm values for primary = 1 and primary = 0
  unique_chr_arm_1 <- setdiff(subset_data$Chr_Arm[subset_data$primary == 1], subset_data$Chr_Arm[subset_data$primary == 0])
  unique_chr_arm_0 <- setdiff(subset_data$Chr_Arm[subset_data$primary == 0], subset_data$Chr_Arm[subset_data$primary == 1])
  
# Check if new rows need to be created
  if (length(unique_chr_arm_1) > 0) {
    new_rows_1 <- data.frame(
      identifier = identifier_value,
      primary = 0,
      Chr_Arm = unique_chr_arm_1,
      pct_affected = 0
    )
  } else {
    new_rows_1 <- NULL
  }
  
  if (length(unique_chr_arm_0) > 0) {
    new_rows_0 <- data.frame(
      identifier = identifier_value,
      primary = 1,
      Chr_Arm = unique_chr_arm_0,
      pct_affected = 0
    )
  } else {
    new_rows_0 <- NULL
  }
  
  # Combine new rows with the original data
  methyl_pr_loss_mod <- bind_rows(methyl_pr_loss_mod, subset_data, new_rows_1, new_rows_0)
}


methyl_pr_gain_mod <- data.frame()

# Loop through identifier values from PR1 to PR13
for (i in 1:3) {
  identifier_value <- paste0("PR", i)
  
  # Filter the data for the current identifier
  subset_data <- methyl_pr_gain %>% filter(identifier == identifier_value)
  
  # Find unique Chr_Arm values for primary = 1 and primary = 0
  unique_chr_arm_1 <- setdiff(subset_data$Chr_Arm[subset_data$primary == 1], subset_data$Chr_Arm[subset_data$primary == 0])
  unique_chr_arm_0 <- setdiff(subset_data$Chr_Arm[subset_data$primary == 0], subset_data$Chr_Arm[subset_data$primary == 1])
  
# Check if new rows need to be created
  if (length(unique_chr_arm_1) > 0) {
    new_rows_1 <- data.frame(
      identifier = identifier_value,
      primary = 0,
      Chr_Arm = unique_chr_arm_1,
      pct_affected = 0
    )
  } else {
    new_rows_1 <- NULL
  }
  
  if (length(unique_chr_arm_0) > 0) {
    new_rows_0 <- data.frame(
      identifier = identifier_value,
      primary = 1,
      Chr_Arm = unique_chr_arm_0,
      pct_affected = 0
    )
  } else {
    new_rows_0 <- NULL
  }
  
  # Combine new rows with the original data
  methyl_pr_gain_mod <- bind_rows(methyl_pr_gain_mod, subset_data, new_rows_1, new_rows_0)
}

```


```{r}
# Prepare for plotting
# Create a mapping from PR1 to PR3 to N to P
identifier_mapping <- setNames(LETTERS[14:16], paste0("PR", 1:3))

# Rename the identifier column
methyl_pr_loss_mod$identifier <- identifier_mapping[methyl_pr_loss_mod$identifier]
methyl_pr_gain_mod$identifier <- identifier_mapping[methyl_pr_gain_mod$identifier]

```


```{r}
# Plotting losses and gains

arm_plot = "1p"   # Adjust the condition accordingly

# Subset the data
subset_data <- methyl_pr_loss_mod %>% filter(Chr_Arm == arm_plot)  

# Create a scatter plot with different colors
scatter_plot_loss <- ggplot(subset_data, aes(x = identifier, y = pct_affected, color = factor(primary, levels = c("1", "0")))) +
  geom_jitter(position = position_dodge(width = 0.5), size = 3) +
  scale_color_manual(values = c("0" = "darkblue", "1" = "lightblue")) +
  labs(y = "Arm Loss (%)") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black"), axis.ticks = element_blank(), axis.text = element_text(size=18), axis.title = element_text(size = 18), legend.position = "none") + scale_y_continuous(breaks = seq(0, 100, by = 25), limits = c(-1,101)) + scale_x_discrete(position = "top") + xlab(NULL)


# Subset the data
subset_data <- methyl_pr_gain_mod %>% filter(Chr_Arm == arm_plot)

# Create a scatter plot with different colors
scatter_plot_gain <- ggplot(subset_data, aes(x = identifier, y = pct_affected, color = factor(primary, levels = c("1", "0")))) +
  geom_jitter(position = position_dodge(width = 0.5), size = 3) +
  scale_color_manual(values = c("0" = "darkred", "1" = "lightcoral")) +
  labs(x = "Patients",
       y = "Arm Gain (%)") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black"), axis.ticks = element_blank(), axis.text = element_text(size=18), axis.title = element_text(size = 18), legend.position = "none") + scale_y_continuous(breaks = seq(0, 100, by = 5), limits = c(-1,11))


# Reverse the y-axis in scatter_plot_loss
scatter_plot_loss_reversed <- scatter_plot_loss + scale_y_reverse(breaks = seq(0, 100, by = 25), limits = c(101, -1)) #scale_y_reverse() #

# Combine the two plots, with scatter_plot_gain on top and scatter_plot_loss_reversed below it
combined_plot <- (scatter_plot_gain / scatter_plot_loss_reversed) +
  plot_layout(heights = c(1, 2))

combined_plot
```